Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-05 index.qmd
section:issues,CaseV; process:write

<!-- 3. But what are limitations of Case V (and the BTL model)? -->
However, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp. 677]. As a result, its assumptions may not be suitable for applications beyond the psycho-physical contexts for which it was created. Thurstone himself cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], acknowledging that some assumptions could prove problematic in the presence of complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376]. Consequently, given that modern CJ applications frequently involve these traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice: the equal dispersion and zero correlation between stimuli.

@@

<!-- 5. But what are the challenges when these assumptions are violated? -->
However, assuming equal dispersions between stimuli despite its violation can lead Case V (and the BTL model) to overlook critical differences in the reliability of the trait across stimuli. Furthermore, if researchers acknowledge that misfit statistics help identify these critical differences, the common practice in the CJ literature of excluding stimuli based on these statistics [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018] risks discarding valuable information and introducing bias into the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases are unpredictable, as they depend on the particular stimuli researchers exclude from the analysis. Taken together, these oversights undermine the reliability of the trait and ultimately compromise its validity [@Perron_et_al_2015, pp. 2].

@@

<!-- 6. So, how do we mitigate these risks? -->
[how do we mitigate these risks?]{style="color:red;"}

<!-- Although no CJ literature offer solutions to mitigate the aforementioned issues, we can take lesson from the Structural Equation Model (SEM) literature. More specifically, from @Depaoli_2021. -->

@@

<!-- 7. What does the zero correlation between stimuli imply? and does it hold? -->
On the other hand hand, the assumption of *zero correlation between stimuli* implies that, during a pairwise comparison, a judge's perception of quality in one text does not influence his perception of the same trait in another text (see @sec-thurstone_theory). Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp. 268]. @Andrich_1978 mathematically demonstrated this cancellation using the BTL model under the assumption of discriminal processes with additive biases. However, it is easy to imagine at least two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli and when an additional hierarchical structure is relevant to the stimuli.

@@

<!-- 8. The first scenario where the assumption does not hold -->
In the first scenario, the intricate aspects of multidimensional, complex traits may introduce dependencies between the stimuli due to certain judges' biases that resist cancellation. Research on text quality suggests that when judges evaluate these traits, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. These additional relevant characteristics, which are unlikely to be equally weighted or opposing, can unevenly influence judges' perceptions, creating biases in their judgments and, ultimately, introducing dependencies between stimuli [@vanderLinden_et_al_2017_II, pp. 346]. For example, this could occur when a judge assessing the argumentative quality of a text places disproportionate emphasis on grammatical accuracy, ultimately favoring texts with fewer errors but weaker arguments. While direct evidence for this specific scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the idea that the factors influencing pairwise comparisons may not always cancel out.

@@

<!-- 10. But what are the challenges when this assumption is violated? -->
In any case, the psychometric and statistical literature emphasizes the importance of addressing the factors that create dependencies between stimuli, as failing to do so can lead to inaccurate conclusions about a trait's reliability and, by extension, its validity. For instance, neglecting additional traits relevant to the stimuli, such as judges' biases, often leads to a dimensional mismatch in the statistical model used for analysis. This mismatch can potentially inflate the reliability of the trait [@Hoyle_et_al_2023, pp. 341] or, worse, introduce bias into the trait's estimates [@Ackerman_1989]. Similarly, failing to account for hierarchical (grouping) structures reduces the precision of model parameter estimates, further amplifying the overestimation of reliability [@Hoyle_et_al_2023, pp. 482]. These issues collectively undermine the trait's reliability and ultimately compromise the validity of the trait's estimates [@Perron_et_al_2015, pp. 2].

@@

<!-- 1. The BTL model as a measurement model for estimating latent traits -->
Building on the previous section, it is evident that, besides its inconveniences, the BTL model commonly functions as the trait's measurement model for CJ data [@Andrich_1978; @Bramley_2008]. A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010]. For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

<!-- 2. Utilizing BTL Scores for further analysis and hypothesis testing -->
Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis tests. For example, these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

@@

<!-- 4. So, how do we mitigate these risks? -->
To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and Item Response Theory (IRT) [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24] recommend conducting these analyses and tests within a structural model. A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010]. This approach allows analyses that can account for both the BTL scores and their uncertainties simultaneously, rather than treating them as separate elements.

>>>>

<!-- 3. But what are limitations of Case V (and the BTL model)? -->
Nevertheless, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp. 677]. As a result, its assumptions may not be suitable for applications beyond the psycho-physical contexts for which it was created. Thurstone himself cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], acknowledging that some assumptions could prove problematic in the presence of complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376]. Consequently, given that modern CJ applications frequently involve these traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice: the equal dispersion and zero correlation between stimuli.

@@

<!-- 5. But what are the challenges when these assumptions are violated? -->
Nonetheless, assuming equal dispersions between stimuli despite its violation can lead Case V (and the BTL model) to overlook critical differences in the reliability of the trait across stimuli, resulting in inaccurate conclusions about the trait's estimates [@McElreath_2020, pp. 370]. Furthermore, if researchers acknowledge that misfit statistics help identify these critical differences, the common practice in the CJ literature of excluding stimuli based on these statistics [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018] risks discarding valuable information and introducing bias into the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases are unpredictable, as they depend on the specific stimuli researchers exclude from the analysis. Taken together, these oversights undermine the reliability of the trait and ultimately compromise its validity [@Perron_et_al_2015, pp. 2].

@@

<!-- 6. So, how do we mitigate these risks? -->
Fortunately, the statistical literature offers a solution to address the abovementioned issues. This solution involves using models that extend traditional approaches to account for the different variability in the stimuli, referred to as over-dispersed models [@McElreath_2020, chap. 12].

@@

<!-- 7. What does the zero correlation between stimuli imply? and does it hold? -->
On the other hand, the assumption of *zero correlation between stimuli* implies that, during a pairwise comparison, a judge's perception of quality in one text does not influence his perception of the same trait in another text (see @sec-thurstone_theory). Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp. 268]. @Andrich_1978 mathematically demonstrated this cancellation using the BTL model under the assumption of discriminal processes with additive biases. However, it is easy to imagine at least two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli and when an additional hierarchical structure is relevant to the stimuli.

@@

<!-- 8. The first scenario where the assumption does not hold -->
In the first scenario, the intricate aspects of multidimensional, complex traits may introduce dependencies between the stimuli due to certain judges' biases that resist cancellation. Research on text quality suggests that when judges evaluate these traits, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. These additional relevant characteristics, which are unlikely to be equally weighted or opposing, can unevenly influence judges' perceptions, creating biases in their judgments and, ultimately, introducing dependencies between stimuli [@vanderLinden_et_al_2017_II, pp. 346]. For example, this could occur when a judge assessing the argumentative quality of a text places more weight on grammatical accuracy than other judges, ultimately favoring texts with fewer errors but weaker arguments. While direct evidence for this specific scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the idea that the factors influencing pairwise comparisons may not always cancel out.

@@

<!-- 10. But what are the challenges when this assumption is violated? -->
In any case, the psychometric and statistical literature emphasizes the importance of addressing the factors that create dependencies between stimuli, as failing to do so can lead to inaccurate conclusions about a trait's reliability and, by extension, its validity. For instance, neglecting additional traits relevant to the stimuli, such as judges' biases, often leads to a dimensional mismatch in the statistical model used for analysis. This mismatch can potentially inflate the reliability of the trait [@Hoyle_et_al_2023, pp. 341] or, worse, introduce bias into the trait's estimates [@Ackerman_1989]. Similarly, failing to account for hierarchical (grouping) structures reduces the precision of model parameter estimates, further amplifying the overestimation of reliability [@Hoyle_et_al_2023, pp. 482]. These issues collectively also undermine the trait's reliability and ultimately compromise the validity of the trait's estimates [@Perron_et_al_2015, pp. 2].

@@

<!-- 1. The BTL model as a measurement model for estimating latent traits -->
Building on the previous section, it is evident that, besides its inconveniences, the BTL model commonly functions as the trait's measurement model in CJ assessments. A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010]. For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

<!-- 2. Utilizing BTL Scores for further analysis and hypothesis testing -->
Researchers then typically use these estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis tests. For example, these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

@@

<!-- 4. So, how do we mitigate these risks? -->
To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and IRT [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24] recommend conducting these analyses and tests within a structural model. A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010]. This approach allows analyses that can account for both the BTL scores and their uncertainties simultaneously, rather than treating them as separate elements.
